### 2022-01-24 plos_latex_template.tex
Version sent to coauthors (w changes by @jananiravi, @andreasancheztapia, @rociojoo)

\newcommand{\fede}[1]{\textcolor{ForestGreen}{Fede: #1}}
\newcommand{\rocio}[1]{\textcolor{MidnightBlue}{Rocio: #1}}
\newcommand{\doro}[1]{\textcolor{RedViolet}{Dorothea: #1}}
\newcommand{\jani}[1]{\textcolor{YellowOrange}{Janina: #1}}
\newcommand{\la}[1]{\textcolor{RawSienna}{Laura: #1}}
\newcommand{\as}[1]{\textcolor{Violet}{Andrea: #1}}
\newcommand{\rita}[1]{\textcolor{CarnationPink}{Rita: #1}}

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Ten simple rules to host an inclusive conference} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Rocío Joo\textsuperscript{1*}, %,2\Yinyang},
Andrea Sánchez-Tapia\textsuperscript{2},
Sara Mortara\textsuperscript{3},
Yanina Bellini Saibene\textsuperscript{4},
Heather Turner\textsuperscript{5}, %\ddag},
Dorothea Hug Peter\textsuperscript{6}, %\ddag},
Matt Bannert\textsuperscript{7},
Batool Almazrouq\textsuperscript{8},
Liz Hare\textsuperscript{9},
Laura Ación\textsuperscript{10},
Juan Pablo Narváez\textsuperscript{11},
Marcela Alfaro Córdoba\textsuperscript{12},
Federico Marini\textsuperscript{13},
Rita Giordano\textsuperscript{14},
Natalia Morandeira\textsuperscript{15},
Adithi Upadhya\textsuperscript{16},
Joselyn Chávez\textsuperscript{17\Yinyang},
Anicet Ebou\textsuperscript{18\Yinyang},
Silvia Canelón\textsuperscript{19\Yinyang},
Janani Ravi\textsuperscript{20}
%,3\textcurrency}
\\
\bigskip
\textbf{1} Global Fishing Watch, Washington, DC 20036, USA
\\
\textbf{2} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{3} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{4} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{5} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{6} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{7} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{8} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{9} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{10} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{11} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{12} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{13} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{14} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{15} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{16} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{17} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{18} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{19} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{20} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\bigskip

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
% 
% Remove or comment out the author notes below if they aren't used.
%
% Primary Equal Contribution Note
\Yinyang These authors contributed equally to this work.

@@

The authors of this article recently organized useR! 2021, the annual global R conference, attracting a broad range of participants from academia, industry, government, and the non-profit sector. To embrace the constraints posed by the pandemic, we hosted a 100\% virtual conference. We worked on building a high-quality virtual and global conference by creating a kind, inclusive, accessible, and welcoming environment for everyone. 
This article outlines our most important lessons learned abstracted as `10 simple rules to host an inclusive conference'. These rules apply equally to academic, industry, or mixed conferences; while inspired by a global experience, these rules seamlessly translate to regional events. 
We have designed them for wide adoption to enable and ensure that the scientific and technical conferences are as accessible and inclusive as they can be.

@@

% Rocío: Main message of the paragraph: Conferences are great, but for some
Conferences are spaces to meet and reconnect with members from a specific community, learn about advances in the field, and share our recent contributions.
A good conference experience can be 
However, opportunities for participating in conferences are not equal for all. 
Many academic and tech conferences have become spaces that reproduce systemic inequalities and privileges, i.e. advantages given by society to people from some groups, like being white, male, from a rich country, English-native speaker, with no physical disabilities \cite{arendDisparityConferenceRegistration2019, biggsAcademicConferenceChilly2018, depickerRethinkingInclusionDisability2020a, irishIncreasingParticipationUsing2020}.
% However, conferences are likely to reproduce the systematic discrimination occurring in other spaces in our fields.

This article proposes 10 simple rules to host inclusive conferences.
They are organized in three sections: \textbf{the pillars}, fundamental rules with key elements to conceive the work on diversity and inclusion in any conference (embracing diversity, creating a welcoming environment, and having a diverse organizing team); \textbf{design rules}, concerning diverse elements in the design of the conference and operational tasks (counteracting bias in the conference program, embracing a strong online component, achieving accessibility for people with disabilities, embracing language diversity, enforcing inclusive communication strategies, and allocating financial resources to support inclusive practices); and a \textbf{process rule} to make the conference part of a long-term commitment to inclusion. 
These practices stem from the experience of organizing useR! 2021, a virtual and global statistical computing conference for users and developers of the R programming language \cite{r_core_team_2021}.
We embraced the challenge of organizing a high-quality virtual conference in the context of the COVID-19 pandemic and making it a kind, inclusive, and accessible experience for as many people as possible.

The authors of this article include the global coordinators of the conference, members of the diversity and inclusion team, and people who contributed to other areas of the conference. We come from different backgrounds, career stages and even continents.
The rules result from our lessons learned from diverse perspectives and expertise. 
Even though these rules were designed based on our experience with a global event, they can be easily transferred or scaled to events with regional or local scopes.
These rules will greatly benefit people who are part of meeting committees that oversee the site/location selection process, or that coordinate with the local organizers of conferences.
The rules will also help organizers who intend to put inclusion at the core of their conference right from the conception and planning phases. We have designed them for wide adoption to enable and ensure that the scientific and technical conferences are as accessible and inclusive as they can be.

% This message is not new. 
% Barriers such as unattainable registration costs, sexism, prejudice against disable people (ableism), among others, have already been exposed and discussed in the literature \cite{arendDisparityConferenceRegistration2019, biggsAcademicConferenceChilly2018, depickerRethinkingInclusionDisability2020a, irishIncreasingParticipationUsing2020}, 
% %timperleyHeMoanaPukepuke2020, gewinWhatScientistsShould2019, brownAbleismAcademiaWhere2018, marks2021meeting
% and some proposals for more inclusive conferences have been put into practice \cite{gichoraTenSimpleRules2010a, levitisCenteringInclusivityDesign2021, atkinsonJournalMedicine20202021, foramittiVirtuesVirtualConferences2021, ninerBetterWhomLeveling2021, rabyMovingAcademicConferences2021, noauthor_discover2021},
% with a primary focus on the online format to open doors for inclusion.\fede{do we need maybe a short definition of the concepts of diversity-equity-inclusion for us/refer to something?} %LA: https://www.cscce.org/resources/organizing-community-events/
% %By inclusion, we mean creating conferences organized by and dedicated to a group of attendees that is representative of the whole population. This means creating a welcoming, respectful, and safe conference that tries to counteract systemic oppression.

% % Rocío: Main message of the paragraph: Roadmap of the paper
% This article focuses on practices for hosting inclusive conferences.
% The rules written here are directed to people who are part of a stable meetings committee that oversees the site/location selection process, or that coordinates with the local organizers of conferences.
% The rules can also be helpful to local/virtual organizers who intend to put inclusion at the core of the planning phase.
% These tips \rita{tips looks to me too informal for an article, maybe used the word `suggestions` or `advice`?} stem from the authors' experience of organizing useR! 2021, a virtual and global statistical computing conference for users and developers of the R programming language \cite{r_core_team_2021}. 
% We embraced the challenge of organizing a high-quality virtual conference in the context of the COVID-19 pandemic and making it a kind, inclusive, and accessible experience for as many persons as possible. 
% Here, we share the lessons learned within the time organizing useR! 2021, summarized as ten simple rules towards hosting an inclusive conference.
% The rules are organized in three groups (Figure \ref{fig:diagram}). %original source: https://docs.google.com/drawings/d/1iS1pLc9OldLMe_jhNT6OGIbT9s5KvlHpUoDd19vHkCY/edit?usp=sharing
% %sara: still need to better format figure size and get the alt text right :P

% Group 1 includes rules 1, 2, and 10, which refer to pillars of an inclusive conference: embracing diversity in all its dimensions, creating a safe and welcoming environment for everyone, and making the conference part of a long-term process for inclusion.
% The second group includes rules 3 and 4 and focuses on the people who participate in the conference. 
% Rule 3 refers to the importance of working with an inclusive and diverse organizing team, and Rule 4 concerns the necessity of counteracting implicit and systemic bias from spotlight roles like keynote speakers, other presenters, program committee members, or other session chairs. 
% The third group includes rules 5 to 9. These rules are about components of the conference that should be carefully planned for: an online component, accessibility for people with disabilities, language inclusiveness, a welcoming communication strategy, and financial resources to support inclusion. 
% These rules apply equally to conferences where the expected audience is academic, from the industry, or a mixed group. While these rules are inspired in a global virtual experience they also apply at the regional or local level and can inform in-person and hybrid events.
% \fede{maybe something like: "Even if this set of rules is derived from our experience with a global scale event, we believe these can be easily transferred to events with regional or local scopes"}

\begin{figure}[!h]
\centering
%sara: without success i'm trying to use pdfcomment package and pdftooltip to make the alt-text. any help?
%ast: R/ Unfortunately, support for adding accessibility metadata to LaTeX documents is limited. You will probably need to use Adobe Acrobat to add missing accessibility metadata to your PDF file.https://chi2021.acm.org/for-authors/presenting/papers/guide-to-an-accessible-submission#latex %ast: let's try not to forget this when submitting - some edits to the PDF may be needed to add alt-text to the final version
\pdftooltip{
\includegraphics[width=\textwidth]{figs/rules.png}}{
%sara: suggestion for alt text:
%Diagram of how the 10 rules are organized into three groups: grounding rules (rules 1, 2, and 10), people-related rules (rules 3 and 4), and design rules (rules 5 to 9). The diagram has five rows. Each line contains rectangular boxes with the rule number and a short title for each rule. Each rule box is colored based on their group (grounding rules are grey, people-related rules are green and design rules are pink). First row is for the grounding rules 1. Multidimensional diversity and 2. Safe and inclusive environment in grey boxes. Second row is for people-related rules 3. Organizing team and 4. Unbias in green boxes. Third row is for the design rules 5. Online component, 6. Accessibility for disabilities, 7. Language inclusivity in pink boxes. The fourth row is for the design rules 8. Communication and 9. Financial support and budget in pink boxes. The last row is for the grounding rule 10. Diversity and inclusion as a process in grey boxes.
}
\caption{Diagram of the rules organized in three groups: the pillars (rules 1, 2, and 3), design rules (rules 4 to 9), and the process (rule 10).}
\label{fig:diagram}
\end{figure}

\section*{I. The pillars}

To organize a more inclusive conference, it is necessary to identify upfront which groups of people are being excluded, the barriers that keep them from active partaking, and strategies to overcome them. 
\textbf{Rule 1} is about embracing all dimensions of diversity and prioritizing the participation of groups that are often marginalized.
However, increasing diversity alone is not enough. \textbf{Rule 2} focuses on how to create a safe and welcoming environment for all the conference attendees. 
Inclusion should start from the organizing team. \textbf{Rule 3} highlights the importance of starting with an inclusive and diverse organizing team, and provides tips on work dynamics.

\subsection*{Rule 1: Embrace all dimensions of diversity}
\label{rule_diversity}

% Rocío: Main idea: what is diversity and why does it matter? (inequalities)
Diversity encompasses multiple dimensions: age, physical ability, career stage, gender, gender identity, geographic origin, language, neurodiversity, race, religion, sexual orientation, and socioeconomic background, to name a few.
Human diversity should be celebrated and respected in every way. 
Unfortunately, we live in a world with implicit hierarchies along these axes. 
Some statuses (e.g. cisgender, white, male, from North America or Western Europe) hold the privilege of being the defaults around which all systems—including conferences—are consciously and unconsciously built. 
People outside the dominant groups suffer different forms of oppression, like sexism, racism, ableism, homophobia, and transphobia; some of them have to deal with several of these on a daily basis.
% While no isolated initiative can drastically alter xx, building a more diverse and inclusive conference starts by recognizing that these inequalities have systematically excluded whole groups of people from academia and scientific and professional circles. 
Therefore, recognizing this systematic exclusion, particularly in our fields and in our scientific or professional communities, is paramount to inclusion, and will help identify the most marginalized and discriminated subgroups \cite{timperleyHeMoanaPukepuke2020}.
% These are the groups that warrant deliberate efforts that ensure fair inclusion.
As a starting point, the meeting organizers could identify which groups warrant deliberate efforts to ensure fair inclusion, and determine the contributing factors (e.g. explicit discrimination, participation fees).
This analysis should guide the vision of diversity and inclusion for your conference and help design strategies to achieve it. 
Express this vision for your conference in a diversity statement. 
Next, use this vision and the set strategies to define measurable indicators and goals of diversity and inclusion (e.g. a gender distribution of your speakers that is representative of the general population, the presence of racialized people (e.g. Black people) among the organizers, speakers, and attendees, or participation from key geographic regions). 
These numbers should guide you and help achieve your vision.

% realistic but bold (back?)
% be aware and promote awareness
% Rocío: more effort towards including the more excluded
% This may translate into different actions depending on your field, region, or community

@@

% Rocío: main idea: Inclusion for everyone
While it is common to think about representation in terms of some of the most visible dimensions of human diversity, such as race, gender, and country of origin, building a genuinely inclusive environment means taking care of all aspects of diversity. %Inclusivity should lie at the heart of setting the tone and meeting space for any meeting as we move forward in the 21st century!
% Edit and make more specific (below)
%\emoji{bulb} \textbf{Practical tips:} 
Here are a few ways in which conference organizers can actively enforce inclusiveness, such as i) Promote respect of pronouns, ii) schedule events exclusive for diverse groups (e.g. LGBTQIA+-friendly spaces), iii) if the conference has an in-person component, pre-assign spaces for religious practices, lactation, and child care, iv) include menu options that account for diverse dietary and religious requirements. These are just a few easy-to-implement examples with a significant impact on including key sub-groups that otherwise find it challenging (see \cite{noauthor_discover2021} for some great advice on `low-hanging fruits').
Importantly, by being proactive in creating such a welcoming space, the organizers can be respectful of attendees' privacy and avoid requiring anyone to disclose personal information, such as revealing a disability, sexual orientation, gender identity, or mental health issues, to name a few.

The organizers should promote a supportive and respectful environment, where all attendants can participate without fear of rejection or isolation. This includes promoting the idea that anyone can act in defense of other attendants in case anything unpleasant, incorrect, of offensive happens. Instead of looking away, act as "active bystanders" and intervene to keep the community safe.
% Rocío: main idea: Code of conduct for a safe place %i think the transition here should leave the two concepts closer. not however. 
However, the most important step that organizers should take to ensure the safety of the community is
Adopting a Code of Conduct (CoC) and gathering a diverse team to enforce it \cite{favaroYourScienceConference2016}.
The CoC is a living document meant to keep the community safe and should state clearly the unacceptable behaviors, the consequences for engaging in such behavior, and the way to report violations \cite{auroraHowRespondCode2019}. 
The CoC team must receive training on how to receive reports, respond to incidents, communicate their responses, and organize protocols on how to respond correspondingly. This should be done during the early phases of the conference planning because CoC violations can happen before the conference too (e.g. social media, organization meetings).  
Since people from marginalized groups are more likely to be victims of Code of Conduct violations and find their claims dismissed due to inadvertent power dynamics, assembling a diverse CoC team will be essential to appropriately deal with discrimination and harassment issues and the power dynamics underneath them with respect and sensitivity. A well-drafted CoC should account for  different cultures and geographical and cultural considerations. 
We strongly recommend reading `How to Respond to Code of Conduct Reports' \cite{auroraHowRespondCode2019} as an excellent starting point for building a strong CoC team.

@@

% Rocío: Idea of this paragraph: build a diverse team, and representation

A genuinely inclusive conference can only be organized by an inclusive and diverse team in charge of the conference decision-making.
Organizers should be a team with people from different regions, genders, ethnicities, socioeconomic statuses, and other aspects of diversity.
Having people from diverse backgrounds in decision-making roles will positively affect the conference as a whole because all the processes will benefit from their expertise, experience, and distinct perspectives \cite{hongGroupsDiverseProblem2004}. 
In addition, a diverse team plays an essential role in creating a welcoming space because representation—seeing people with similar life experiences occupy critical positions of power and breaking negative stereotypes—is one of the best ways to create a sense of belonging for everyone participating in the conference (see \textbf{Rule 2} about creating a safe and welcoming environment).

% Rocío: Idea of this paragraph: non-transferability and do not restrict people
People with disabilities often say: `Nothing about us without us' [ref]; the same holds for other dimensions of diversity. The experience of people in marginalized groups facing barriers for equal participation and trying to solve them are invaluable; they cannot be replaced by good intentions or second-hand knowledge from people who have not lived through the same experiences \cite{costanzachockDesign2020}.
% People who have experienced exclusion have the social and technical expertise to fight against it and they will be essential in any team.%ast maybe frame this on a more positive note (ideas, problem solving)
It is important, though, to ensure that organizers with diverse backgrounds are not restricted to only work on diversity and inclusion aspects; 
every member should be granted the freedom to choose which areas of the conference they want to work on, in addition to contributing to decisions involving marginalized communities.

% Rocío: Idea of this paragraph: tips to make a diverse and inclusive team work + paying them

%\emoji{bulb} \textbf{Practical tips:} 
If you have already started assembling an organizing team, check for gaps in its composition and find representatives accordingly. 
The regional and local communities, groups, or associations in your field are good sources to tap into. 
Furthermore, self-nomination and voting do not always work as reliable mechanisms to fill leadership positions. Instead, it helps to nominate members directly and offer them leading positions in the organization team, especially those that would be typically occupied by people from privileged groups.
Build an environment that encourages healthy conflict and constructive criticism, one in which every person can express their opinion and pay special attention to people from systematically excluded groups.
Many people lack the institutional support to put time and effort into the organization tasks and do not have the luxury to commit to the organization for free; consider paying them as an item in your budget as well as other types of recognition (e.g. public recognition on social media, on the conference website, or during the conference, ideally whatever works best for the contributing member).
Furthermore, tasks such as receiving and responding to Code of Conduct reports can be emotionally intense work and should be additionally rewarded.

% % Rocío: might go to the next rule
% Splitting the workload and responsibilities should not be done by putting care-taking labors--community building, meeting organization and note-taking, conversations with potential partners--on the hands of women and other minoritized groups, while people from privileged groups take the lead in stereotypical highly-valued tasks (see \textbf{Rule \ref{rule_unbias}}).

\section*{II. Design Rules}

>>>>

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Ten simple rules to host an inclusive conference} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Rocío Joo\textsuperscript{1*\Yinyang}, 
Andrea Sánchez-Tapia\textsuperscript{2\Yinyang},
Sara Mortara\textsuperscript{3},
Yanina Bellini Saibene\textsuperscript{4},
Heather Turner\textsuperscript{5}, %\ddag},
Dorothea Hug Peter\textsuperscript{6}, %\ddag},
Matt Bannert\textsuperscript{7},
Batool Almazrouq\textsuperscript{8},
Liz Hare\textsuperscript{9},
Laura Ación\textsuperscript{10},
Juan Pablo Narváez-Gómez\textsuperscript{11},
Marcela Alfaro Córdoba\textsuperscript{12},
Federico Marini\textsuperscript{13},
Rita Giordano\textsuperscript{14},
Natalia Morandeira\textsuperscript{15},
Adithi Upadhya\textsuperscript{16},
Joselyn Chávez\textsuperscript{17},
Anicet Ebou\textsuperscript{18},
Silvia Canelón\textsuperscript{19},
Janani Ravi\textsuperscript{20}
%,3\textcurrency}
\\
\bigskip
\textbf{1} Global Fishing Watch, Washington, DC 20036, USA
\\
\textbf{2} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{3} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{4} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{5} Department of Statistics, University of Warwick, Coventry, UK
\\
\textbf{6} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{7} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{8} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{9} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{10} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{11} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{12} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{13} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{14} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{15} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{16} ILK Labs/Bengaluru, India - 560046
\\
\textbf{17} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{18} Bioinformatic team, Département de Formation et de Recherches en Agriculture et Ressources Animales, Institut National Polytechnique Félix Houphouët-Boigny, BP 1313 Yamoussoukro, Côte d’Ivoire.
\\
\textbf{19} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{20} Departments of Pathobiology and Diagnostic Investigation, Microbiology and Molecular Genetics, Michigan State University, East Lansing, MI 48824, USA
\\
\bigskip

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
% 
% Remove or comment out the author notes below if they aren't used.
%
% Primary Equal Contribution Note
\Yinyang These authors share first authorship of this work.

@@

Conferences are spaces to meet and network within an academic or technical field, to learn about new advances and share our work. They can help define career paths and create long-lasting collaborations and opportunities. 
However, these opportunities are not equal for all. 
This article introduces 10 simple rules to host an inclusive conference. 
The authors recently organized the 2021 edition of the useR! statistical computing conference, attracting a broad range of participants from academia, industry, government, and the non-profit sector. 
Coming from different backgrounds, career stages and even continents, we embraced the challenge of organizing a high-quality virtual conference in the context of the COVID-19 pandemic and making it a kind, inclusive, and accessible experience for as many people as possible.
The rules result from our lessons learned before, during, and after the organization of the conference. 
They have been mainly written for potential organizers or selection committees of conferences, and contain multiple practical tips so that they can help diverse types of scientific and technological conferences to become as accessible and inclusive as they can and should be.

@@

% Conferences are great, but only for some
Conferences are spaces to meet and reconnect with members from a specific community, learn about advances in the field, and share our recent contributions.
A good conference experience can make a difference in the professional development of the participants and create long-lasting collaborations and opportunities. 
However, opportunities for participating in conferences are not equal for all. 
Many academic and tech conferences have become spaces that reproduce systemic inequalities, by failing to eliminate the barriers for participation and giving more opportunities to the most privileged individuals (typically white people from rich countries, prestigious institutions, native English speakers with no physical disabilities, for example) \cite{arendDisparityConferenceRegistration2019, biggsAcademicConferenceChilly2018, depickerRethinkingInclusionDisability2020a, irishIncreasingParticipationUsing2020}.
The authors of this article have experienced these inequalities in conferences, from different perspectives and levels of inclusion/exclusion and privilege.
We had the opportunity to organize a virtual and global statistical computing conference, useR! 2021 \cite{sanchez-tapia_user_2021-2}, for users and developers of the R programming language \cite{r_core_team_2021}.
Coming from different backgrounds, career stages and even continents, we embraced the challenge of organizing a high-quality virtual conference in the context of the COVID-19 pandemic and making it a kind, inclusive, and accessible experience for as many people as possible.

% Roadmap of the paper
The rules result from our lessons learned before, during, and after the organization of the useR! conference. 
They are organized in three sections: the pillars, design rules, and the continuity rule (\textbf{Fig. \ref{fig:diagram}}).
\textbf{The pillars} comprise three fundamental rules with key elements to conceive the work on diversity and inclusion in any conference. 
\textbf{Rule 1} is about setting a vision of diversity and inclusion that should guide all the efforts and decision making in the organization.
\textbf{Rule 2} focuses on how to create a safe and welcoming environment for all the attendees. 
\textbf{Rule 3} highlights the importance of starting with an inclusive and diverse organizing team and provides tips on work dynamics.
Six \textbf{design rules} focus on weaving inclusion into the conference design process.
In \textbf{Rule 4}, we introduce multiple ways to counteract bias in the conference program (keynotes, program committee, abstract selection, and other spaces). 
\textbf{Rule 5} provides advice for organizing inclusive online and hybrid conferences.
\textbf{Rule 6} focuses on accessibility practices to include people with disabilities. 
In \textbf{Rule 7}, we provide suggestions to encourage and ensure full participation of non-native English speakers. 
\textbf{Rule 8} provides tips for developing an inclusive communication strategy. 
In \textbf{Rule 9}, we address budgeting for inclusive practices and helping participants with affordable registration costs, scholarships, and other forms of financial support.
Finally, 
a \textbf{continuity Rule} to make the conference part of a long-term commitment to inclusion.

We believe that these rules will greatly benefit organizers who intend to put inclusion at the core of their conference right from the conception and planning phases.
The rules will also help people who are part of meeting committees that oversee the site/location selection process, or that coordinate with the local organizers of conferences. 
We have designed them for wide adoption to help scientific and technical conferences to become as accessible and inclusive as they can and should be.

% refs for past works in case we add them in line 311.
% %timperleyHeMoanaPukepuke2020, gewinWhatScientistsShould2019, brownAbleismAcademiaWhere2018, marks2021meeting
% some proposals for more inclusive conferences have been put into practice \cite{gichoraTenSimpleRules2010a, levitisCenteringInclusivityDesign2021, atkinsonJournalMedicine20202021, foramittiVirtuesVirtualConferences2021, ninerBetterWhomLeveling2021, rabyMovingAcademicConferences2021, numfocus_discover_2021},

\begin{figure}[!h]
\centering
%ast: use Adobe Acrobat to add missing accessibility metadata to your PDF file. let's try not to forget this when submitting - some edits to the PDF may be needed to add alt-text to the final version https://chi2021.acm.org/for-authors/presenting/papers/guide-to-an-accessible-submission#late
\pdftooltip{
\includegraphics[width=\textwidth]{figs/rules.png}}{
%Alt-text for the figure to be included in the final PDF: Diagram of how the 10 rules are organized into three groups: pillar rules (rules 1-3), design rules (rules 4 to 9), and a continuity rule (rule 10). The overall diagram depicts an arrow from left to right which contains all the rules. From left to right, "the pillars" label is surrounded by blue bubbles which show Rule "1 Vision", Rule "2 Safe environment", and Rule "3 Organizing Team". Then, the label "Design Rules" is surrounded by six blue bubbles with Rules "4 Counteract bias", "5 Online", "6 Accessibility", "7 Language", "8 Communication",  "9 Financial". To the right, and forming the head of the arrow, the Tenth Rule is under the label "Continuity" and consists of a blue bubble "10 Assess and pass along".
}
\caption{Schematic diagram of the rules organized in three groups: the pillars (Rules 1, 2, and 3), design rules (Rules 4 to 9), and continuity (Rule 10).}
\label{fig:diagram}
\end{figure}

\subsection*{Rule 1: Set a vision for diversity and inclusion}
\label{rule_diversity}

Diversity encompasses multiple dimensions: age, physical ability, career stage, gender, gender identity, geographic origin, language, neurodiversity, race, religion, sexual orientation, and socioeconomic background, to name a few.
Human diversity should be celebrated and respected in every way. 
Nonetheless, we live in a world with implicit hierarchies along these axes. 
Some statuses (e.g. being cisgender, white, male, from North America or Western Europe) hold the privilege of being the defaults around which all systems—including conferences—are consciously and unconsciously built. 
People outside the dominant groups suffer different forms of oppression, like sexism, racism, ableism, homophobia, and transphobia, and sometimes these forms of oppression happen simultaneously and in complex ways.
As a result, some groups of people have been systematically excluded from or only partially included in academia and scientific and professional circles.
Recognizing this systematic exclusion is paramount to inclusion.
Assess how it translates to your particular field and scientific or professional community and use this assessment to set the vision and goals of diversity and inclusion for your conference.
Define measurable indicators for these goals (e.g. a gender distribution of your speakers that is representative of the general population, the participation of people from diverse races and ethnicities among the organizers, speakers, and attendees, or participation from key geographic regions) that will help you be accountable along the way.
The vision can be expressed in a diversity statement to be used as a reference for decision making in all areas of the conference organization.

@@

% Inclusion for everyone
% Explain why we need this
Real inclusion can only be achieved in an environment where everybody feels safe, respected, and invited to take part of the conference activities and community. Be proactive and design a welcoming conference experience that takes into account the well-being of all the attendees. This means respecting all the aspects you committed to in your diversity vision (see \textbf{Rule 1}). For example, promote the use and respect of preferred pronouns and commit to respect all genders and sexual orientations. If the conference has an in-person component, pre-assign quiet, private spaces for religious practices, lactation, and child care. Include menu options that account for diverse dietary and religious requirements.
Some of these actions are very easy to implement and have a significant impact on making everyone feel truly included (see \cite{numfocus_discover_2021} for more examples and  great advice on `low-hanging fruits').
Open the schedule for events organized by thematic groups and communities (e.g. taskforce meetings, LGBTQIA+-friendly spaces, Black in STEM), and sessions aimed to welcome first-timers to the conference.

% A culture of support (active bystandership) and code of conduct for a safe place 
Creating a safe and welcoming environment also requires promoting a culture of support, including the idea that anyone can intervene respectfully in defense of other attendees in case anything unpleasant, incorrect, or offensive happens.
This is called `active bystandership'; consider providing training or pre-conference discussions to promote it.
However, the most important step that you should take to ensure the safety of the community is
adopting a code of conduct (coc) and preparing a diverse response team to enforce it \cite{favaroYourScienceConference2016}.
The coc is a living document meant to keep the community safe and should state clearly the unacceptable behaviors, the consequences for engaging in such behaviors, and the way to report violations \cite{auroraHowRespondCode2019}.
The coc response team must receive training on how to receive reports, respond to incidents, and communicate their responses to the whole community. The team should be prepared early during conference planning because coc violations can happen before the conference too (e.g. social media, organization meetings).  
We strongly recommend reading `How to Respond to code of conduct Reports' \cite{auroraHowRespondCode2019} as an excellent starting point for building a strong coc team.

@@

% Build a diverse team, and representation
A genuinely inclusive conference can only be organized by an inclusive and diverse team.
Organizers should comprise a team with people from as many different backgrounds as possible regarding regions, genders, ethnicities, career stages, and other aspects of diversity.
People with disabilities often say: `Nothing about us without us' \cite{charlton_nothing_1998, werner_nothing_1998}; the same holds for other dimensions of diversity. 
Having people from diverse backgrounds in decision-making roles will positively affect the conference as a whole because all the processes will benefit from their expertise, experience, and distinct perspectives \cite{hongGroupsDiverseProblem2004}. 
The experience of people in marginalized groups
is especially important; they cannot be replaced by good intentions or second-hand knowledge from people who have not lived through the same experiences \cite{costanzachockDesign2020}.
In addition, a diverse team plays an essential role in creating a welcoming space because representation—seeing people with similar life experiences occupy critical positions of power and breaking negative stereotypes—is one of the best ways to create a sense of belonging for everyone participating in the conference.
It is important, though, to ensure that organizers with diverse backgrounds are not restricted to only work on diversity and inclusion aspects; 
every member should have the freedom to choose which areas of the conference they want to contribute to.

% Rocío: Idea of this paragraph: tips to make a diverse and inclusive team work + paying them
If you have already started assembling an organizing team, check for gaps in its composition and do your best to fill them. 
The regional and local communities, groups, or associations in your field are good sources to tap into. 
When it comes to team leadership, do not expect self-nomination and voting to work as infallible mechanisms to fill these positions equitably. 
People from underrepresented groups sometimes refrain to nominate themselves while people in dominant groups tend to do it more easily. 
Instead, nominate folks directly and offer them leading positions in the organization team, especially those positions that would be typically occupied by people from privileged groups.
Bear in mind that some people may lack the institutional or financial support to put time and effort into the organization tasks and do not have the luxury to commit to the organization for free; consider paying them as an item in your budget (see \textbf{Rule 9}).
Furthermore, tasks such as receiving and responding to code of conduct reports can be emotionally intense work and should be additionally rewarded.
And most importantly, take care of your team and their well-being! Check on them regularly and make sure that everyone is comfortable. This represents a great amount of work and you will need to support each other in the long run.

\subsection*{Rule 4: Consciously counteract bias in the conference program}
\label{rule_unbias}

The conference program may lack diversity in the invited roles––such as keynote speakers, program/scientific committee, session chairs, panelists––, the accepted abstracts, and even the scope of the topics covered. 
It is important to counteract biases in all of these aspects.
When choosing or inviting people for visible and valued roles in the conference it is likely that there will not be much diversity in the first pool of names \cite{dwyerNoticeWhoScience2021,swartzScienceValueDiversity2019,wongBuildDiversityScience2020,dignazioUnicornsJanitorsNinjas2020}. 
Rather than deter you, this should encourage you to go beyond your current networks to look for, reach out to, invite, encourage, and onboard great people that are not routinely in the spotlight, making sure that these roles do not reinforce existing privileges (e.g. `manels' and all-white panels \cite{else_how_2019}).

Biases in the final list of abstracts begin with self-selection: some potential conference participants may not feel confident enough to submit their work, especially to large and prestigious conferences.
To mitigate this self-selection bias, 
some communities have pre-submission mechanisms to aid their members prepare and receive feedback for their abstract (e.g. the Latin-R and R-Ladies communities have volunteers to help their members improve their conference abstracts).
If this does not exist in your field yet, consider asking members of your community to provide a similar kind of mentorship.
